\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-14 ext-ind-type-compilation-example.tex
compilation of family STLC

\begin{lstlisting}
(* Type Checking of Inductive Type *) [@\tanc{Tyck-IndTy-start}@]
(* Tyck happens during interactively creating families*)
Module Type STLC_Ctx1. 
(* The Context including Inductive Type *)
  Parameter (tm : Set).
  Parameter (tm_var : id → tm).
  Parameter (tm_abs : id → tm → tm).
  (* [@\dadada@] and all the fields for inductive type [@\dadada@] *)
End STLC_Ctx1.
Module Type tm_prec_64 (self: STLC_Ctx1).[@\YZ{tm_6? Do you mean STLC_Ctx1 instead?}\EDJreply{Yes Thanks.}@]
Parameter (* Type for partial recursor *)
(tm_prec : ∀ t, ∀ P : self.tm -> Type,
  (∀ n, option (P (self.tm_var n))) -> 
  [@\dadada@], option (P t)).
End tm_prec_64.
Module Type STLC_Ctx2.  
  Include STLC_Ctx1.
  Include tm_prec_64. (* New Ctx include prec *)
End STLC_Ctx2.
(* [@\dadada@] etc, everytime a new field is added, a new 
module type is created as the ctx with the new field *)
(* Collect eliminator TYPE for later type-check *)
Module tm_rec_type (self: STLC_Ctx1).
  Definition _TYPE_tm_rec :=
    ∀ P : self.tm -> Set,
    ( ∀ n, P (self.tm_var n)) ->
    ( ∀ n i, P i -> P (self.tm_abs n i)) ->
    [@\dadada@] -> ∀ i : self.tm, P i.
End tm_rec_type.  [@\tanc{Tyck-IndTy-end}@]
\end{lstlisting}

@@

\begin{lstlisting}
(* Type checking of Recursion *) [@\tanc{Tyck-RecTy-start}@]
(* Typecheck each handler, here STLC.subst.tm_var *)
(* Also an example of fields in nested family [@\dadada@]*) 
Module tm_var_handle (STLC: STLC_Ctx1).[@\YZ{(1) Why doesn't the module name contain "subst"? (2) Should it be STLC_Ctx2 instead of STLC_Ctx1?}\EDJreply{I rename to ctx so that I think it is easier to understand. Feel Free to change to the way you want. I think name them all into "_Ctx" suffix is easier to emphasize the fact they are the result of compilation of the context. (2) The fact is, no, it also will need to include tm_prec's computational axiom and ty, which I also forgot to mention. Also ty has its prec and computational axiom. So at least 6. Do you want me to write doesn STLC_Ctx(1,2,3,4,5,6)? It is unnecessary complication and the space is not enough. I proposed to move this part of information into the figure 5 toy example in the slack, what do you think? }@]
[@\YZ{(3) Are nested families really needed?}\EDJreply{Not really but if consider handlers as one family, it is nested family. What's more, the audience needs to know how nested family is compiled at some point.}@]
  Definition tm_var s x t := 
    if eqb x s then t else STLC.tm_var s.
End tm_var_handle. (* [@\dadada@] other handlers are alike *)
(* Assembling tycked handlers in STLC.subst *)
Module subst_internal_410 (M: STLC_Ctx1). 
 Module subst_internal.
 Include tm_var_handle M. Include tm_abs_handle M.
 Include tm_app_handle M. Include tm_unit_handle M.
 End subst_internal.
End subst_internal_410.
(* Intermediate Module solely for exhausitivity checking *) 
Module do_tc (self: STLC_Ctx[@\dadada@]).
  Include tm_rec_type self. 
  Include subst_internal_410 self.
  Parameter (for_tyck : _TYPE_tm_rec).
  Definition term_for_type_checking :=
    for_tyck  ( λ _  ⇒ (id → self.tm → self.tm))
      subst_internal.tm_var [@\dadada@] .
End do_tc.[@\tanc{Tyck-RecTy-end}@]
\end{lstlisting}

@@

\begin{lstlisting}
(* Exposing Computational Axiom into Ctx *) [@\tanc{Tyck-Recty-cont-start}@]
Module Type substtm_139 (self: STLC_Ctx[@\dadada@]).[@\YZ{What is "..."?}\EDJreply{My intension is STLC_Ctx_n for some n where at this point the audience should able to see what the context of that STLC_Ctx is. Apparently I failed.}@]
Parameter
  (subst_on_tm_unit :
	 (self.subst (self.tm_unit)) =[@\YZ{Where is this "subst" added to the context?}\EDJreply{The reason is above. Currently the space is already not enough.}@]
      self.subst_internal.tm_unit).
(* [@\dadada@] subst equality on other constructors *)
End substtm_139. [@\tanc{Tyck-Recty-cont-end}@]
\end{lstlisting}

@@

\begin{lstlisting}
(* Compilation, happens when closing a top-level family *) [@\tanc{Cmp-Result-start}@]
(* Ctx that expose concrete inductive type *)
Module Type STLC_CCtx.
Inductive __internal_tm : Set := [@\dadada@] 
Include subst_internal_410. (* [@\dadada@] and more inclusion *)
End STLC_CCtx.
(* Both the compilation of recursor and[@\dadada@] *)
Module subst_128 (self: STLC_CCtx).
Include __motive_of_subst_107 self.
Definition subst :=
  self.__internal_tm_rec __motiveTsubst
	self.subst_internal.tm_var [@\dadada@] .
End subst_128.
(* [@\dadada@]the compilation of partial recursor 
require concrete inductive type exposed in the ctx *)
Module tm_prec_72 (self: STLC_CCtx[@\dadada@]).
Definition tm_prec :
  ∀ t, ∀ P : self.tm -> Type,
  (∀ n, option (P (self.tm_var n))) ->
  [@\dadada@], option (P t) := [@\dadada@]
End tm_prec_72. 
Module STLC. (* Assembling/Compiling the whole STLC *)
Include tm_5. Include tm_prec_72. 
Include tm_prectm_81. (*Justify comp axiom, mostly eq_refl*)
Include subst_internal_410. Include subst_128.
End v_129.
(* Assembling STLC_bool.subst_internal via reusing 
    most of the former compilation results *)
(* Notice how handlers don't require explicitly
   exposing the inductive type *)
Module subst_internal_248 (self: STLC_Ctx1).
(* The New Handlers *)
Include tm_true_197 self. Include [@\dadada@]
(* The Reused Compilation result *)
Include tm_var_handle self. Include [@\dadada@]
End v_248. 
(* Regenerated FRecursion Compilation *)
Module subst_242 (self: [@\dadada@]).
Include __motive_of_subst_107 self.
Definition subst :=
  self.__internal_tm_rec __motiveTsubst
	self.subst_internal.tm_var [@\dadada@]
    self.subst_internal.tm_true [@\dadada@] .
End subst_242. [@\tanc{Cmp-Result-end}@]
\end{lstlisting}

>>>>

\begin{lstlisting}
(* Type Checking of Inductive Type *) [@\tanc{Tyck-IndTy-start}@]
(* Tyck happens during interactively creating families*)
Module Type STLC_Ctx1. 
(* The Context including Inductive Type *)
  Parameter (tm : Set).
  Parameter (tm_var : id → tm).
  Parameter (tm_abs : id → tm → tm).
  (* [@\dadada@] and all the fields for inductive type [@\dadada@] *)
End STLC_Ctx1.
Module Type tm_prec_64 (self: STLC_Ctx1).
Parameter (* Type for partial recursor *)
(tm_prec : ∀ t, ∀ P : self.tm -> Type,
  (∀ n, option (P (self.tm_var n))) -> 
  [@\dadada@], option (P t)).
End tm_prec_64.
Module Type STLC_Ctx2.  
  Include STLC_Ctx1.
  Include tm_prec_64. (* New Ctx include prec *)
End STLC_Ctx2.
(* [@\dadada@] etc, everytime a new field is added, a new 
module type is created as the ctx with the new field *)
(* Collect eliminator TYPE for later type-check *)
Module tm_rec_type (self: STLC_Ctx1).
  Definition _TYPE_tm_rec :=
    ∀ P : self.tm -> Set,
    ( ∀ n, P (self.tm_var n)) ->
    ( ∀ n i, P i -> P (self.tm_abs n i)) ->
    [@\dadada@] -> ∀ i : self.tm, P i.
End tm_rec_type.  [@\tanc{Tyck-IndTy-end}@]
\end{lstlisting}

@@

\begin{lstlisting}
(* Type checking of Recursion *) [@\tanc{Tyck-RecTy-start}@]
(* Typecheck each handler, here STLC.subst.tm_var *)
(* Also an example of fields in nested family [@\dadada@]*) 
Module tm_var_handle (STLC: STLC_Ctx1).[@\YZ{(1) Why doesn't the module name contain "subst"? (2) Should it be STLC_Ctx2 instead of STLC_Ctx1?}\EDJreply{I rename to ctx so that I think it is easier to understand. Feel Free to change to the way you want. I think name them all into "_Ctx" suffix is easier to emphasize the fact they are the result of compilation of the context. (2) The fact is, no, it also will need to include tm_prec's computational axiom and ty, which I also forgot to mention. Also ty has its prec and computational axiom. So at least 6. Do you want me to write doesn STLC_Ctx(1,2,3,4,5,6)? It is unnecessary complication and the space is not enough. I proposed to move this part of information into the figure 5 toy example in the slack, what do you think? }@]
  Definition tm_var s x t := 
    if eqb x s then t else STLC.tm_var s.
End tm_var_handle. (* [@\dadada@] other handlers are alike *)
(* Assembling tycked handlers in STLC.subst *)
Module subst_internal_410 (M: STLC_Ctx1). 
 Module subst_internal.
 Include tm_var_handle M. Include tm_abs_handle M.
 Include tm_app_handle M. Include tm_unit_handle M.
 End subst_internal.
End subst_internal_410.
(* Intermediate Module solely for exhausitivity checking *) 
Module do_tc (self: STLC_Ctx[@\dadada@]).
  Include tm_rec_type self. 
  Include subst_internal_410 self.
  Parameter (for_tyck : _TYPE_tm_rec).
  Definition term_for_type_checking :=
    for_tyck  ( λ _  ⇒ (id → self.tm → self.tm))
      subst_internal.tm_var [@\dadada@] .
End do_tc.[@\tanc{Tyck-RecTy-end}@]
\end{lstlisting}

@@

\begin{lstlisting}
(* Exposing Computational Axiom into Ctx *) [@\tanc{Tyck-Recty-cont-start}@]
Module Type substtm_139 (self: STLC_Ctx[@\dadada@]).
Parameter
  (subst_on_tm_unit :
	 (self.subst (self.tm_unit)) =
      self.subst_internal.tm_unit).
(* [@\dadada@] subst equality on other constructors *)
End substtm_139. [@\tanc{Tyck-Recty-cont-end}@]
\end{lstlisting}

@@

\begin{lstlisting}
(* Compilation, happens when closing a top-level family *) [@\tanc{Cmp-Result-start}@]
(* Ctx that exposes concrete inductive type *)
Module Type STLC_CCtx.
Inductive __internal_tm : Set := [@\dadada@] 
Include subst_internal_410. (* [@\dadada@] and more inclusion *)
End STLC_CCtx.
(* Both the compilation of recursor and[@\dadada@] *)
Module subst_128 (self: STLC_CCtx).
Include __motive_of_subst_107 self.
Definition subst :=
  self.__internal_tm_rec __motiveTsubst
	self.subst_internal.tm_var [@\dadada@] .
End subst_128.
(* [@\dadada@]the compilation of partial recursor 
require concrete inductive type exposed in the ctx *)
Module tm_prec_72 (self: STLC_CCtx[@\dadada@]).
Definition tm_prec :
  ∀ t, ∀ P : self.tm -> Type,
  (∀ n, option (P (self.tm_var n))) ->
  [@\dadada@], option (P t) := [@\dadada@]
End tm_prec_72. 
Module STLC. (* Assembling/Compiling the whole STLC *)
Include tm_5. Include tm_prec_72. 
Include tm_prectm_81. (*Justify comp axiom, mostly eq_refl*)
Include subst_internal_410. Include subst_128.
End v_129.
(* Assembling STLC_bool.subst_internal via reusing 
    most of the former compilation results *)
(* Notice how handlers don't require explicitly
   exposing the inductive type *)
Module subst_internal_248 (self: STLC_Ctx1).
(* The New Handlers *)
Include tm_true_197 self. Include [@\dadada@]
(* The Reused Compilation result *)
Include tm_var_handle self. Include [@\dadada@]
End v_248. 
(* Regenerated FRecursion Compilation *)
Module subst_242 (self: [@\dadada@]).
Include __motive_of_subst_107 self.
Definition subst :=
  self.__internal_tm_rec __motiveTsubst
	self.subst_internal.tm_var [@\dadada@]
    self.subst_internal.tm_true [@\dadada@] .
End subst_242. [@\tanc{Cmp-Result-end}@]
\end{lstlisting}

### 2022-10-14 lang-design.tex
compilation of family STLC

A key observation is that although late binding
prevents \emph{definitional equality} on \lsti{subst}, it does not
affect \emph{propositional equality}.
That is, \lsti{subst tm_unit x t' = tm_unit}, as a proposition
(\lsti{Prop}), should still hold.
After all, how \lsti{subst} is defined on \lsti{tm_unit} does not vary
from a base family to a derived family.
%
Based on this insight, our plugin automatically generates
propositional equalities for each recursive function defined with \lsti{FRecursion},
making them available as axioms for use by the rest of the current family.
The plugin also provides a tactic \lsti{fsimpl} that enables, for instance,
simplifying \lsti{subst tm_unit x t'} to \lsti{tm_unit}.
Under the hood, \lsti{fsimpl} works by rewriting using the postulated
propositional equalities.

@@

\begin{centered}
\begin{minipage}{.96\textwidth}
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
STLC_tm_prec : \forall (P : tm -> Set), [@\hlcase{tm-partialrec-case-var}@](\forall x, option (P (tm_var x)))[@\tikzmarkend{tm-partialrec-case-var}@] -> [@\hlcase{tm-partialrec-case-unit}@]option (P tm_unit)[@\tikzmarkend{tm-partialrec-case-unit}@] ->
               [@\hlcase{tm-partialrec-case-abs}@](\forall x t, option (P t) -> option (P (tm_abs x t)))[@\tikzmarkend{tm-partialrec-case-abs}@] ->
               [@\hlcase{tm-partialrec-case-app}@](\forall t1, option (P t1) -> \forall t2, option (P t2) -> option (P (tm_app t1 t2)))[@\tikzmarkend{tm-partialrec-case-app}@] ->
               \forall t, option (P t)
\end{lstlisting}
\end{minipage}
\end{centered}

\noindent
As expected, "STLC_tm_prec" is axiomatized along with
four (propositional) equalities, one for each constructor of "tm".
For instance, the equality for the "tm_abs" constructor is as follows:

\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
\forall x t P H1 H2 H3 H4,
STLC_tm_prec P H1 H2 H3 H4 (tm_abs x t) = H3 x t (STLC_tm_prec P H1 H2 H3 H4 t)
\end{lstlisting}

%\noindent
Importantly, unlike the standard "tm_rec" recursor, the partial recursor
"STLC_tm_prec" is compatible with the late binding of "tm" in its type.
When "STLC_tm_prec" is inherited into family "STLCFix", all the previous four equalities
still hold, and a trivial, fifth equality is made available:

\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
\forall x t P H1 H2 H3 H4, STLC_tm_prec P H1 H2 H3 H4 (tm_fix x t) = None
\end{lstlisting}

Although partial recursors appear weaker than ordinary recursors,
there is power in restraint.
Partial recursors offer a principled, uniform way to derive injectivity and disjointness of
constructors, while allowing extensibility:
they enable injective mappings from values of a late bound inductive
type, like "tm", to values of an ordinary inductive type, like \lsti{\nat},
the injectivity and disjointness of whose constructors are readily
available.
%
For example, in a proof state with the assumption \lsti{H : tm_var "x" = tm_abs "y" t},
running \lsti{fdiscriminate H} first uses \lsti{H} to obtain the equality

\begin{centered}
\begin{minipage}{\textwidth}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
STLC_tm_prec (\lambda[@\,@]_, \nat) (\lambda[@\,@]_, Some 1) (Some 2) (\lambda[@\,@]_[@\,@]_[@\,@]_, Some 3) (\lambda[@\,@]_[@\,@]_[@\,@]_[@\,@]_, Some 4) (tm_var "x") =
STLC_tm_prec (\lambda[@\,@]_, \nat) (\lambda[@\,@]_, Some 1) (Some 2) (\lambda[@\,@]_[@\,@]_[@\,@]_, Some 3) (\lambda[@\,@]_[@\,@]_[@\,@]_[@\,@]_, Some 4) (tm_abs "y" t)
\end{lstlisting}
\end{minipage}
\end{centered}

\noindent
and then uses the axiomatized equalities about "STLC_tm_prec" to
rewrite the equality above, obtaining \lsti{Some 1 = Some 4},
from which \lsti{False} easily follows.

Note that the proof term generated by \lsti{fdiscriminate H} in family "STLC"
is reusable by family "STLCFix", because the partial recursor and its
equalities are compatible with the late binding of "tm".
Also note that within family "STLCFix", a second partial recursor
(called "STLCFix_tm_prec") on "tm" is available;
it can be used to derive properties about "tm_fix",
such as \lsti{tm_var "x" = tm_fix "y" t -> False}.

\noindentparagraph{Limitations.}

>>>>

A key observation is that although late binding
prevents \emph{definitional equality} on \lsti{subst}, it does not
affect \emph{propositional equality}.
That is, \lsti{\forall x t, subst tm_unit x t = tm_unit},
as a proposition (\lsti{Prop}), should still hold.
After all, how \lsti{subst} is defined on \lsti{tm_unit} does not vary
from a base family to a derived family.
%
Based on this insight, our plugin automatically generates
propositional equalities for each recursive function defined with \lsti{FRecursion},
making them available as axioms for use by the rest of the current family.
The plugin also provides a tactic \lsti{fsimpl} that enables, for instance,
simplifying \lsti{subst tm_unit x t'} to \lsti{tm_unit}.
Under the hood, \lsti{fsimpl} works by rewriting using the postulated
propositional equalities.

@@

\begin{centered}
\begin{minipage}{.96\textwidth}
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
tm_prec_STLC : \forall (P : tm -> Set), [@\hlcase{tm-partialrec-case-var}@](\forall x, option (P (tm_var x)))[@\tikzmarkend{tm-partialrec-case-var}@] -> [@\hlcase{tm-partialrec-case-unit}@]option (P tm_unit)[@\tikzmarkend{tm-partialrec-case-unit}@] ->
               [@\hlcase{tm-partialrec-case-abs}@](\forall x t, option (P t) -> option (P (tm_abs x t)))[@\tikzmarkend{tm-partialrec-case-abs}@] ->
               [@\hlcase{tm-partialrec-case-app}@](\forall t1, option (P t1) -> \forall t2, option (P t2) -> option (P (tm_app t1 t2)))[@\tikzmarkend{tm-partialrec-case-app}@] ->
               \forall t, option (P t)
\end{lstlisting}
\end{minipage}
\end{centered}

\noindent
As expected, "tm_prec_STLC" is axiomatized along with
four (propositional) equalities, one for each constructor of "tm".
For instance, the equality for constructor "tm_abs" is as follows:

\begin{centered}
\begin{minipage}{.99\textwidth}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
tm_abs_eq_STLC: \forall x t P H1 H2 H3 H4,
                tm_prec_STLC P H1 H2 H3 H4 (tm_abs x t) = H3 x t (tm_prec_STLC P H1 H2 H3 H4 t)
\end{lstlisting}
\end{minipage}
\end{centered}

%\noindent
Importantly, unlike the standard "tm_rec" recursor, the partial recursor
"tm_prec_STLC" is compatible with the late binding of "tm" in its type.
When "tm_prec_STLC" is inherited into family "STLCFix", all the previous four equalities
still hold, and a trivial, fifth equality is made available:

\begin{centered}
\begin{minipage}{.88\textwidth}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
tm_fix_eq_STLC : \forall x t P H1 H2 H3 H4, tm_prec_STLC P H1 H2 H3 H4 (tm_fix x t) = None
\end{lstlisting}
\end{minipage}
\end{centered}

Partial recursors appear weaker than ordinary recursors,
but there is power in restraint.
Partial recursors offer a principled, uniform way to derive injectivity and disjointness of
constructors, while allowing for extensibility:
they enable injective mappings from a late bound inductive
type, like "tm", to an ordinary inductive type, like \lsti{\nat},
the injectivity and disjointness of whose constructors are readily
available.
%
For example, in a proof state with the assumption \lsti{H : tm_var "x" = tm_abs "y" t},
running \lsti{fdiscriminate H} first applies the injective mapping to
both sides of~\lsti{H}, obtaining

\begin{centered}
\begin{minipage}{.96\textwidth}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
tm_prec_STLC (\lambda[@\,@]_, \nat) (\lambda[@\,@]_, Some 1) (Some 2) (\lambda[@\,@]_[@\,@]_[@\,@]_, Some 3) (\lambda[@\,@]_[@\,@]_[@\,@]_[@\,@]_, Some 4) (tm_var "x") =
tm_prec_STLC (\lambda[@\,@]_, \nat) (\lambda[@\,@]_, Some 1) (Some 2) (\lambda[@\,@]_[@\,@]_[@\,@]_, Some 3) (\lambda[@\,@]_[@\,@]_[@\,@]_[@\,@]_, Some 4) (tm_abs "y" t)
\end{lstlisting}
\end{minipage}
\end{centered}

\noindent
and then rewrites the above equality using two of the axiomatized
equalities about "tm_prec_STLC",
%(namely "tm_var_eq_STLC" and "tm_abs_eq_STLC"),
obtaining \lsti{Some 1 = Some 4},
from which \lsti{False} easily follows.

Note that the proof term generated by \lsti{fdiscriminate H} in "STLC"
is reusable by family "STLCFix", because the partial recursor and its
equalities are compatible with the late binding of "tm".
Also note that within family "STLCFix", a second partial recursor
(called "tm_prec_STLCFix") and its equalities are available,
allowing properties about "tm_fix",
such as \lsti{\neg(tm_var "x" = tm_fix "y" t)}, to be proved.

%\noindentparagraph{Limitations.}

\ifShowOldWriting
